Supporting Information: Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-02-19 paper/rebuttal/rebuttal.tex

We find that, using the best performing dimensions (300) and window size (1),  $\gamma = 1.0$, and for default experimental parameters (global mobility, population defined using all mobile and non-mobile scholars), that the embedding distance, or cosine distance between vectors learned through neural embedding, explains more of the flux, and leads to better predictions than other baselines (Fig.~\ref{fig:supp:distancemetrics}).
This performance held when considering both the exponential and the power-law forms of the gravity model of mobility.
The euclidean distance between directly-factorized vectors performs comparably, and even beats out the embedding distance under certain experimental conditions (Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).

\subsection{Why do neural embeddings outperform direct matrix factorization?}% why sgd is better?
Why do neural-embedding approaches, which rely on stochastic gradient descent, outperform direct matrix factorization, especially given that \textit{word2vec} \emph{is} implicitly approximating factorization?
An issue in the matrix factorization approach is that it is extremely sensitive to small flows between affiliations. The matrix factorization approach embeds affiliations $i$ and $j$ such that their dot similarity is as close as to $\log (T_{ij} / P(i) P(j))$. If the flow $T_{ij}$ is considerably small or zero, the dot similarity goes $-\infty$, pushing $i$ and $j$ very far from the other affiliations in the embedding space~\autocite{levy2014neural, Qui2018}. This is particularly problematic when the window size is small because most affiliation pairs would have no flow, which is indeed the case in our experiments. To circumvent this problem, the previous studies~\autocite{levy2014neural, Qui2018} add a constant flow between the affiliation pairs with no flow. However, this changes the data to which we fit the model, and small flows still have a considerably strong impact on the embedding.

Also, it is well known that SVD itself is vulnerable to outliers \autocite{xu2012robust, huber1981robust, xu1995robust, chandrasekaran2011rank, candes2011robust}.
The stochastic gradient descent algorithm (SGD), which is employed in SGNS \textit{word2vec}, is more robust than SVD and can enhance generalization and effectiveness of {\it word2vec} model \autocite{ma2018power, smith2020generalization, zhang2019algorithmic}.

### 2021-02-19 paper/current/supp.tex
add speculation para

\paragraph*{S8 Text}
\label{si:text:levy}
{\bf Why do neural embedings outperform direct matrix factorization?}

Why do neural-embedding approaches, which rely on stochastic gradient descent, outperform direct matrix factorization, especially seeing as \textit{word2vec} implicitly approximates factorization?
An issue in the matrix factorization approach is that it is extremely sensitive to small flows between affiliations. The matrix factorization approach embeds affiliations $i$ and $j$ such that their dot similarity is as close as to $\log (T_{ij} / P(i) P(j))$. If the flow $T_{ij}$ is considerably small or zero, the dot similarity goes $-\infty$, pushing $i$ and $j$ very far from the other affiliations in the embedding space~\autocite{levy2014neural, Qui2018}. This is particularly problematic when the window size is small because most affiliation pairs would have no flow, which is indeed the case in our experiments. To circumvent this problem, the previous studies~\autocite{levy2014neural, Qui2018} add a constant flow between the affiliation pairs with no flow. However, this changes the data to which we fit the model, and small flows still have a considerably strong impact on the embedding.

@@

%
% S8 TEXT
\paragraph*{S8 Text}
\label{si:text:organizations}
{\bf Organization disambiguation and metadata.}

@@

%
% S9 TEXT
\paragraph*{S9 Text}
\label{si:text:disambiguation}
{\bf Author name dismabiguation.}

@@

%
% S10 TEXT
\paragraph*{S10 Text}
\label{si:text:network_ranking}
{\bf Reconstructing Times ranking with network measure.}

@@

%
% S11 TEXT
\paragraph*{S11 Text}
\label{si:text:boomerang_speculation}
{\bf Speculation on variations of the convex-curve pattern.}

>>>>

\paragraph*{S8 Text}
\label{si:text:whynueralbetter}
{\bf Why do neural embedings outperform direct matrix factorization?}

Why do neural-embedding approaches, which rely on stochastic gradient descent, outperform direct matrix factorization, especially given that \textit{word2vec} \emph{is} implicitly approximating factorization?
An issue in the matrix factorization approach is that it is extremely sensitive to small flows between affiliations. The matrix factorization approach embeds affiliations $i$ and $j$ such that their dot similarity is as close as to $\log (T_{ij} / P(i) P(j))$. If the flow $T_{ij}$ is considerably small or zero, the dot similarity goes $-\infty$, pushing $i$ and $j$ very far from the other affiliations in the embedding space~\autocite{levy2014neural, Qui2018}. This is particularly problematic when the window size is small because most affiliation pairs would have no flow, which is indeed the case in our experiments. To circumvent this problem, the previous studies~\autocite{levy2014neural, Qui2018} add a constant flow between the affiliation pairs with no flow. However, this changes the data to which we fit the model, and small flows still have a considerably strong impact on the embedding.

@@

%
% S8 TEXT
\paragraph*{S9 Text}
\label{si:text:organizations}
{\bf Organization disambiguation and metadata.}

@@

%
% S9 TEXT
\paragraph*{S10 Text}
\label{si:text:disambiguation}
{\bf Author name dismabiguation.}

@@

%
% S10 TEXT
\paragraph*{S11 Text}
\label{si:text:network_ranking}
{\bf Reconstructing Times ranking with network measure.}

@@

%
% S11 TEXT
\paragraph*{S12 Text}
\label{si:text:boomerang_speculation}
{\bf Speculation on variations of the convex-curve pattern.}
